Normalization by Evaluation for Sized Dependent Types

### 2017-02-19 nbe/icfp17.tex
[ ICFP 17 ] Reification rules, o as offset, N for Up terms.

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \zero a \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid k \mid \ind i + k ~(k \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.

@@

Size increment \fbox{$a + k$} for $k \in \NN$ extends addition by $\infty + k = \infty$ and $(\ind i + k') + k = \ind i + (k' + k)$.  Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $k \leq k'$ where either $a = k$ and $b = k'$ or $a \in \{ k, \ind i + k \}$ and $b = \ind i + k'$.

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (b + 1) \qquad
      \Gamma \der T : \Nat (b + 1) \to \Set[\ell] \\
      \Gamma \der a_z,a_s : \Size \qquad
      \Gamma \der t_z : T \, (\zero{a_z}) \qquad
      \Gamma \der t_s : (t : \Nat b) \to T \, (\suc{a_s} t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat b \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,b \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

      \Gamma \der a_z,a_s : \Size \qquad
      \Gamma \der t_z : T\,\zero{a_z} \qquad
      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{a_s}n)
    }{\Gamma \der (\zero{a_z})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}

@@

      \Gamma \der a_z,a_s : \Size \qquad
      \Gamma \der t_z : T\,\zero{a_z} \qquad
      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{a_s}n)
    }{\Gamma \der (\suc{a_s} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a_s} t)}

@@

We extend the syntax by de Bruijn levels $\lev k$
to be used as generic values (unknowns) and type annotations
$\up T t$ and $\down T t$ for lazy realisations of the
reflection and reification operations of NbE.
\[
\begin{array}{lllrl@{\qquad}l}
%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
\end{array}
\]

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ a \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ \ann a \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{(\zero a) @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc a t) @ \casett \evalsto w}
\qquad
  \ru{t\,\ann a\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero a, \suc a u \}}
\end{gather*}
The rules added for NbE deal with elimination of delayed reflection:
\begin{gather*}
  \ru{T' \evalsto \piT U T
    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
\qquad
  \ru{T' \evalsto \piT \Size T
    }{(\up {T'} n) @ a \evalsto \up{T[a]} (n \, a)}
\qquad
  \ru{T' \evalsto \forallT T
    }{(\up {T'} n) @ \ann a \evalsto \up{T[a]} (n \ann a)}
%   \ru{T \evalsto \piT A B \qquad
%       \up{B[u]} (n \, \down A u) \evalsto w
%     }{(\up T n) @ u \evalsto w}
% \qquad
%   \ru{T \evalsto \piT \Size B \qquad
%       \up{B[u]} (n \, a) \evalsto w
%     }{(\up T n) @ a \evalsto w}
% \qquad
%   \ru{T \evalsto \forallT B \qquad
%       \up{B[a]} (n \ann a) \evalsto w
%     }{(\up T n) @ \ann a \evalsto w}
\\[2ex]
  \ru{T' \evalsto \Nat b \qquad
    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
      \up{T\,(\up{\Nat \infty} n)} (n\,\case[\ell] {(\down {\Set[\infty]} T)} {(\down {T\,{\zero\infty}} t_z)} {(t_s)})
    }
\end{gather*}

@@

>>>>

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \zero a \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid o \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size \\
\end{array}
\]
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a (named dependent) function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.

@@

Size increment \fbox{$a + o'$} for $o' \in \NN$ extends addition by $\infty + o' = \infty$ and $(\ind i + o) + o' = \ind i + (o + o')$.  Size comparison \fbox{$a \leq b$} holds if either $b = \infty$ or  $o \leq o'$ where either $a = o$ and $b = o'$ or $a \in \{ o, \ind i + o \}$ and $b = \ind i + o'$.

@@

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (a + 1) \qquad
      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, (\zero{a}) \qquad
      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat a \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

      \Gamma \der t_z : T\,\zero{b} \qquad
      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
    }{\Gamma \der (\zero{a})\,\case[\ell] T {t_z} {t_s} = t_z : T\,\zero{a_z}}

@@

      \Gamma \der t_z : T\,\zero{b} \qquad
      \Gamma \der t_s : (n : \Nat b) \to T\,(\suc{b}n)
    }{\Gamma \der (\suc{a} t)\,\case[\ell] T {t_z} {t_s} = t_s\,t : T\,(\suc{a} t)}

@@

We extend the syntax by de Bruijn levels $\lev k$
to be used as generic values (unknowns) and type annotations
$\up T t$ and $\down T t$ for lazy realisations of the
reflection and reification operations of NbE.
\[
\begin{array}{lllrl@{\qquad}l}
%  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
  \Ne   & \ni & n & ::= & \dots \mid \lev k \\
  \tUp  & \ni & N & ::= & \up T n \\
  \Whnf & \ni & w & ::= & \dots \mid N \mid \down T t \\
%  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
\end{array}
\]

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ a \evalsto w}
\qquad
  \ru{t[a] \evalsto w
    }{(\lambda t) @ \ann a \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{(\zero a) @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc a t) @ \casett \evalsto w}
\\[2ex]
  \ru{t\,\ann a\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero a, \suc a u \}}
\end{gather*}
The rules added for NbE deal with elimination of delayed reflection:
\begin{gather*}
  \ru{T' \evalsto \piT U T
    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
\qquad
  \ru{T' \evalsto \piT \Size T
    }{(\up {T'} n) @ a \evalsto \up{T[a]} (n \, a)}
\qquad
  \ru{T' \evalsto \forallT T
    }{(\up {T'} n) @ \ann a \evalsto \up{T[a]} (n \ann a)}
%   \ru{T \evalsto \piT A B \qquad
%       \up{B[u]} (n \, \down A u) \evalsto w
%     }{(\up T n) @ u \evalsto w}
% \qquad
%   \ru{T \evalsto \piT \Size B \qquad
%       \up{B[u]} (n \, a) \evalsto w
%     }{(\up T n) @ a \evalsto w}
% \qquad
%   \ru{T \evalsto \forallT B \qquad
%       \up{B[a]} (n \ann a) \evalsto w
%     }{(\up T n) @ \ann a \evalsto w}
\\[2ex]
  \ru{T' \evalsto \Nat b \qquad
    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
      \up{T\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             T %% {(\down {\Set[\infty]} T)}
             {(\down {T\,{\zero\infty}} t_z)}
             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
    }
\\[2ex]
  \ru{T' \evalsto \Nat b \qquad
    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
    }
\end{gather*}

\subsection{Read back}
\label{sec:read}

Judgements.
\[
\begin{array}{l@{\qquad}l}
   \R k t  v     & \mbox{} \\
   \RTy k T  V  & \mbox{} \\
   \RNat k t  v & \mbox{} \\
   \RNe k t  m  & \mbox{} \\
\end{array}
\]
\fbox{$\R k t v$}
\begin{gather*}
  \ru{t \evalsto \down{U} T \qquad
      U \evalsto s \\
      \RTy k T V
    }{\R k t V}
\qquad
  \ru{t \evalsto \down{U} u \qquad
      U \evalsto \Nat a \\
      \RNat k u v
    }{\R k t v}
\qquad
  \ru{t \evalsto \down{U} u \qquad
      U \evalsto \up {T'} {t'} \\
      \RNe k u v
    }{\R k t v}
\\[2ex]
  \ru{t \evalsto \down{U} f \qquad
      U \evalsto \piT A B \qquad
      \R{k+1} {\down{B[\up A \lev k]} (f\,\up A \lev k)} v
    }{\R k t {\lambda v}}
\\[2ex]
  \ru{t \evalsto \down{U} f \qquad
      U \evalsto \piT \Size B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\lev k)} v
    }{\R k t {\lambda v}}
\qquad
  \ru{t \evalsto \down{U} f \qquad
      U \evalsto \forallT B \qquad
      \R{k+1} {\down{B[\lev k]} (f\,\ann{\lev k})} v
    }{\R k t {\lambda v}}
\end{gather*}

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{t \evalsto n \qquad
      \RNe k n m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{a_v} \qquad
      \RSize k {a_v} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{a_v}u \qquad
      \RSize k {a_v} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

\fbox{$\RSize k {a_v} a$}
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k-1-j} + o}}
\end{gather*}

\fbox{$\RNe k t m$} and \fbox{$\RElim k e e^v$}
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k t {\ind{k-1-j}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k t v
    }{\RElim k t v}
\qquad
  \ru{\RSize k a b
    }{\RElim k a b}
\qquad
  \ru{\RSize k a b
    }{\RElim k {\ann a} {\ann b}}
\\[2ex]
  \ru{\RTy k T V \qquad
      \R k {t_z} {v_z} \qquad
      \R k {t_s} {v_s}
    }{\RElim k {(\case[\ell] T {t_z} {t_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\RTy k T V \qquad
      \R k t v
    }{\RElim k {(\fix[\ell] T t)} {\fix[\ell] V v}}
\end{gather*}
\fbox{$\RTy k T V$}
\begin{gather*}
  \ru{T \evalsto \Set[\ell]
    }{\RTy k T {\Set[\ell]}}
\qquad
  \ru{T \evalsto \Nat a \qquad
      \RSize k a b
    }{\RTy k T {\Nat b}}
\qquad
  \ru{T \evalsto \up{U}u \qquad
      \RNe k u m
    }{\RTy k T m}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      \RTy k A {V_a} \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\piT {V_a} V_b}}
\qquad
  \ru{T \evalsto \epiT \Size B \qquad
      \RTy {k+1} B {V_b}
    }{\RTy k T {\epiT \Size V_b}}
\end{gather*}

@@

\subsection{Partial equivalence relations}
\label{sec:per}

A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive. The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A)$ the relation $\A$ is in fact an equivalence relation.

The PERs $\NE$ and $\NF$ charaterize (neutral) normalizing terms.
\[
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
  \fbox{$t = t' \in \NE$}
    & \defas &
    \RNe k t m &\mand& \RNe k {t'} {m'} &\mand& m \approx m' &\mforall k \\
  \fbox{$t = t' \in \NF$}
    & \defas &
    \R k t v &\mand& \R k {t'} {v'} &\mand& v \approx v' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \approx m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \approx n'
  %   }{t = t' \in \NE}
\]
All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

\subsection{PER model}
\label{sec:permode}

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{T \evalsto N \qquad T' \evalsto N'
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NE
\\[3ex]
  \ru{T \evalsto \Nat\,a \qquad T' \evalsto \Nat\,a
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(a)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; a \mapsto \EL[\ell](B[a]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[a] = B'[a] \in \SET[\ell] \mforall a \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(a \mapsto \EL[\ell](B[a]))
\end{array}
\]
